Gradual Soundness: Lessons from Static Python

### 2022-04-24 pj-7.1/paper.tex
pj: review edits v2

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptX}{\typefont{X}} %% undefined type for an early example
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptstr}{\typefont{Str}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptflout}{\typefont{Flout}}
\newcommand{\sptdyn}{\typefont{Dyn}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptset}[1]{\paramtype{\typefont{Set}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{Dict}}
\newcommand{\sptrawchkdict}{\typefont{CheckedDict}} %% not a real SP type, but useful for TeX
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\sptrawchkdict}{#1, #2}}
\newcommand{\sptenv}{\typefont{Env}}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\trule}[1]{\textsc{#1}}

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\sptclass}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\wtsub}{\vdash}
\newcommand{\spsubt}{\mathrel{<:}}
\newcommand{\spsubteq}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spconsist}{\spcompat}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spstr}{\typefont{str}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[3]{\spapp{\paramtype{\code{chkdict}}{#1, #2}}{#3}}
\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
\newcommand{\spdictref}[2]{#1[#2]}
\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
\newcommand{\spobjref}[2]{#1.#2}
\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}

@@

Generic types are more difficult to enforce because a tag by itself does not
describe their contents.
For example, suppose that a function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for Python dictionaries is not enough to protect the
function against dicts with ill-typed elements.
The \SP{} solution is to provide a second kind of dictionary that ensures
the validity of its elements.
These different values have different types:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a tag check
    for dict values that ignores elements.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    does not promise anything about keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a
    three-part check for concrete dicts with exactly the same key and value
    types.
    The overall cost of such a check depends on the nesting of concrete types,
    but (crucially) does not depend on the size of incoming values.
    %% Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options keep run-time costs low and let programmers decide between
concrete and shallow types.
Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

@@

For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
with keys of type $K$ and values of type $V$.
The implementation has three main components:
\begin{itemize}
  \item
    The constructor requires two type parameters and a Python dictionary ($\sptrawpydict$).
    It checks that the elements of the dictionary match the key and value types.
  \item
    All untyped writes must be protected by casts.
    Every operation that mutates or extends a checked dictionary must
    validate any untyped arguments that it receives.
  \item
    The dictionary value stores a compound type tag 
    (e.g., one token representing a three-part type $\sptchkdict{K}{V}$)
    to support casts from the dynamic type.
    When a concrete dictionary enters typed code from an untyped context,
    its tag must match the expected type exactly.
    For example, type $\sptchkdict{K}{V}$ matches the type
    $\sptchkdict{K'}{V'}$ only if $K$ is equal to $K'$
    and $V$ is equal to $V'$.
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a compound tag.

@@

\Cref{f:surface-types} presents a core syntax for programs.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require type annotations.
Functions must have one positional argument.
Classes must declare one parent (either $\spobject$ or another class), one field, and one method.
Expressions describe values and simple computations.
The basic values are the none value, integers,
booleans (which are the integers \code{0} and \code{1}), strings, and the top object.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
Other expressions describe function calls, dictionary reads and writes, object field reads and writes,
and method calls.

@@

Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We assume that all unions are written in a flat and simplified form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be normalized to $\sptdyn$.

@@

\begin{figure}[t]
  \begin{langarray}
    \spprog & \defeq &
      \spstmt \langmid \spstmt,~\spprog
    \\[1.5ex]
    \spstmt & \defeq &
      \spvardef{\spx}{\sptype}{\spexpr} \langmid
      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
  \\ & &
      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr}}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr}}
    \\[1.5ex]
    \spexpr & \defeq &
      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spstr \langmid \spobject \langmid
  \\ & &
      \sppydict{\spx: \spexpr, \ldots} \langmid
      \spchkdict{\sptype}{\sptype}{\sppydict{\spx: \spexpr, \ldots}} \langmid
  \\ & &
      \spapp{\spf}{\spexpr} \langmid
      \spdictref{\spexpr}{\spexpr} \langmid
      \spdictset{\spexpr}{\spexpr}{\spexpr} \langmid
      \spobjref{\spc}{\spx} \langmid
      \spobjset{\spc}{\spx}{\spexpr} \langmid
      \spobjapp{\spc}{\spf}{\spexpr}
    \\[1.5ex]
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptstr \langmid
      \sptobject \langmid
      \sptclass \langmid
  \\ & &
      \sptpydict{\sptype}{\sptype} \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptunion{\sptype, \ldots}
    \\[1.5ex]
    \sptenv & \defeq &
      \sptenvnil \langmid
      \sptvardef{\spx}{\sptype},~\sptenv \langmid
      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype}}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype}},~\sptenv
    \\[1.5ex]
    \spx, \spf & \defeq & \mbox{variable names}
  \end{langarray}

@@

  \caption{Surface Syntax and Types}
  \label{f:surface-types}
\end{figure}

@@

  \caption{Evaluation Types, Surface-to-Evaluation Mapping}
  \label{f:surface-to-eval-types}
\end{figure}

Evaluation types $\spteval$ are a subset of the surface types.
These are the types that \SP{} promises to soundly enforce.
\Cref{f:surface-to-eval-types} presents the syntax of evaluation types
and a retraction $\mftypeF{\cdot}$ from surface types to evaluation types.
As the retraction shows, the evaluation types make two main changes:
\begin{enumerate}
  \item
    The parameterized type for Python dictionaries
    $\sptpydict{\sptype_0}{\sptype_1}$ gets replaced with
    a raw type $\sptrawpydict$.
    The raw type behaves the same as $\sptpydict{\sptdyn}{\sptdyn}$ would.
  \item
    Unions get replaced with the dynamic type, except for the special case
    of unions with none ($\sptoptional{\sptype_0}$).
\end{enumerate}

@@

\begin{table}[t]
  \caption{How to Enforce the Evaluation Types}
  \label{t:cast}

  \centering
  \(\begin{array}{l@{\quad}l}
    \mbox{Eval. Type $\spteval$} & \mbox{Description of run-time check} \\\hline
    \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
    \sptrawpydict & \mbox{Accepts any Python dictionary} \\
    \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
    \sptdyn & \mbox{No check needed, accepts any value} \\
    \sptobject & \mbox{Accepts any value, except primitives~(\cref{s:c-types})} \\
    \sptclass_0 & \mbox{Accepts instances of the class $\sptclass_0$} \\
    \sptnone & \mbox{Accepts the Python none value} \\
    \sptint & \mbox{Accepts integers} \\
    \sptbool & \mbox{Accepts booleans (the integers \code{0} and \code{1})} \\
    \sptstr & \mbox{Accepts strings}
  \end{array}\)
\end{table}

@@

    \inferrule*{
      \sptclassdef{\sptclass_0}{\sptclass_2}{\ldots}{\ldots} \in \sptenv_0
      \\
      \sptenv_0 \wtsub \sptclass_2 \spsubt \sptclass_1
    }{
      \sptenv_0 \wtsub \sptclass_0 \spsubt \sptclass_1
    }

@@

  \caption{Expression Typing, Consistent Subtyping, and Materialization}
  \label{f:eval-types}
\end{figure}

@@

We end here, with materialization, rather than present a semantics
for the formalization.
After all, the main benefit of a full formal semantics is to validate
the behavior of complex expressions---and this job is
better left to the Redex mechanization.

@@

%\subsubsection{Program Typing, Inheritance Restriction}
%
%\Cref{f:program-typing} presents the typing judgment for programs,
%which folds over a sequence of statements to build a type environment.
%The typing rules for variables and functions are standard.
%We draw attention to the rule for classes, in particular the override condition.
%A class may override fields and methods from its parent;
%if it does, the overriding definition must be a \emph{static subtype} of the original.
%For example, a method that returns an integer can be overridden by a method that returns
%a boolean, but it cannot be overriden by a method that returns the dynamic type.
%This restriction lets the \SP{} compiler avoid some run-time checks; it can assume
%that static methods return reliable values instead of having to insert checks for subtypes.
%
%\begin{figure}[t]
%  \fbox{$\sptenv \wtprog \spprog \dashv \sptenv$}
%  \begin{mathpar}
%    \inferrule*{
%      \sptenv_0 \wtprog \spstmt_0 \dashv \sptenv_1
%      \\\\
%      \sptenv_1 \wtprog \spprog_0 \dashv \sptenv_2
%    }{
%      \sptenv_0 \wtprog \spstmt_0, \spprog_0 \dashv \sptenv_2
%    }
%
%    \inferrule*{
%      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
%    }{
%      \sptenv_0 \wtprog \spvardef{\spx_0}{\spteval_0}{\spexpr_0} \dashv \sptvardef{\spx_0}{\spteval_0},~\sptenv_0
%    }
%
%    \inferrule*{
%      \sptenv_1 = \sptfundef{\spf_0}{\spteval_0}{\spteval_1},~\sptenv_0
%      \\\\
%      \sptenv_1 \wtexpr \spexpr_0 : \spteval_1
%    }{
%      \sptenv_0 \wtprog \spfundef{\spf_0}{\spann{\spx_0}{\spteval_0}}{\spteval_1}{\spexpr_0} \dashv \sptenv_1
%    }
%
%    \inferrule*{
%      \sptenv_1 = \sptclassdef{\spc_0}{\spc_1}{\sptvardef{\spx_0}{\spteval_0}}{\sptfundef{\spf_1}{\spc_0, \spteval_1}{\spteval_2}}, \sptenv_0
%      \\\\
%      %% TODO are field overrides exact?
%      \mbox{if $\spann{\spx_0}{\spteval_0'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\spteval_0 = \spteval_0'$}
%      \\\\
%      %% TODO can T override dyn?
%      \mbox{if $\sptfundef{\spf_1}{\spc_2, \spteval_1'}{\spteval_2'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\spteval_1' \spsubteq \spteval_1$ and $\spteval_2 \spsubteq \spteval_2'$}
%      \\\\
%      \sptenv_1 \wtexpr \spvardef{\spexpr_0} : \spteval_0
%      \\
%      \spann{\spself}{\spc_0}, \spann{\spx_1}{\spteval_1}, \sptenv_1 \wtexpr \spexpr_1 : \spteval_2
%    }{
%      \sptenv_0 \wtprog \spclassdef{\spc_0}{\spc_1}{\spvardef{\spx_0}{\spteval_0}{\spexpr_0}}{\spfundef{\spf_1}{\spself, \spann{\spx_1}{\spteval_1}}{\spteval_2}{\spexpr_1}} \dashv \sptenv_1
%    }
%  \end{mathpar}
%
%  \caption{Program Typing}
%  \label{f:program-typing}
%\end{figure}

>>>>

%% 2022-04-24 More model issues:
%% - C.f(e) uses a class type, not an instance ... how to invoke method on instance?!
%% ... we should really apologize for a too-simple model
%% - Object is not clearly in C
%% - introduce c again as variable name? gee why not expr.method()?

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{S}}
\newcommand{\spteval}{\typefont{T}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptX}{\typefont{X}} %% undefined type for an early example
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptstr}{\typefont{Str}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptflout}{\typefont{Flout}}
\newcommand{\sptdyn}{\typefont{Dyn}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptset}[1]{\paramtype{\typefont{Set}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{Dict}}
\newcommand{\sptrawchkdict}{\typefont{CheckedDict}} %% not a real SP type, but useful for TeX
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\sptrawchkdict}{#1, #2}}
\newcommand{\sptenv}{\typefont{Env}}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\trule}[1]{\textsc{#1}}

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\sptclass}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\wtsub}{\vdash}
\newcommand{\spsubt}{\mathrel{<:}}
\newcommand{\spsubteq}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spconsist}{\spcompat}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spstr}{\codefont{str}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[3]{\spapp{\paramtype{\code{chkdict}}{#1, #2}}{#3}}
\newcommand{\spcnew}[2]{\spapp{#1}{#2}}
\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
\newcommand{\spdictref}[2]{#1[#2]}
\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
\newcommand{\spobjref}[2]{#1.#2}
\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}

@@

Generic types are more difficult to enforce because a tag by itself does not
describe their contents.
For example, suppose that a function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for Python dictionaries is not enough to protect the
function against dicts with ill-typed elements.
The \SP{} solution is to provide a second kind of dictionary that ensures
the validity of its elements.
These different values have different types:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a tag check
    for dict values that ignores elements.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    does not promise anything about keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a
    check for concrete dicts that have exactly the same key and value
    types.
    To implement such checks efficiently, \SP{} stores a run-time registry
    of instantiated concrete types.
    Every concrete dict carries a pointer into this registry to enable
    constant-time type validation, even for nested concrete types.
    %% Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options keep run-time costs low and let programmers decide between
concrete and shallow types.
Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

@@

For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
with keys of type $K$ and values of type $V$.
The implementation has three main components:
\begin{itemize}
  \item
    The constructor requires two type parameters and a Python dictionary ($\sptrawpydict$).
    It checks that the elements of the dictionary match the key and value types.
  \item
    All untyped writes must be protected by casts.
    Every operation that mutates or extends a checked dictionary must
    validate any untyped arguments that it receives.
  \item
    The dictionary value stores a type tag to support casts from the dynamic
    type.
    A tag represents a three-part type $\sptchkdict{K}{V}$ and is implemented
    as a pointer into a global registry of instantiated types.
    When a concrete dictionary enters typed code from an untyped context,
    its tag must match the expected type exactly.
    For example, type $\sptchkdict{K}{V}$ matches the type
    $\sptchkdict{K'}{V'}$ only if $K$ is equal to $K'$
    and $V$ is equal to $V'$.
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a tag.
The tag is supported by a global registry of instantiated concrete types.

@@

\Cref{f:surface-types} presents a core syntax for programs.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require type annotations.
Functions must have one positional argument.
Classes must declare one parent (either $\sptobject$ or another class), one field, and one method.
Expressions describe values and simple computations.
The basic values are the none value, integers,
booleans (which are the integers \code{0} and \code{1}), strings, and the top object.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
Other expressions describe function calls, dictionary reads and writes, object field reads and writes,
and method calls.

@@

Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We choose to write the dynamic type as $\sptdyn$ for simplicity; technically,
\SP{} does not have a surface syntax for its dynamic
type and instead allows missing and unknown types~(\cref{s:type-dynamic}).
We assume that all unions are written in a flat and simplified form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be normalized to $\sptdyn$.

@@

\begin{figure}[t]
  \begin{langarray}
    \spprog & \defeq &
      \spstmt \langmid \spstmt,~\spprog
    \\[1.5ex]
    \spstmt & \defeq &
      \spvardef{\spx}{\sptype}{\spexpr} \langmid
      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
  \\ & &
      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr}}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr}}
    \\[1.5ex]
    \spexpr & \defeq &
      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spstr \langmid
  \\ & &
      \sppydict{\spx: \spexpr, \ldots} \langmid
      \spchkdict{\sptype}{\sptype}{\sppydict{\spx: \spexpr, \ldots}} \langmid
      \spcnew{\sptclass}{\spexpr, \ldots} \langmid
  \\ & &
      \spapp{\spf}{\spexpr} \langmid
      \spdictref{\spexpr}{\spexpr} \langmid
      \spdictset{\spexpr}{\spexpr}{\spexpr} \langmid
      \spobjref{\spc}{\spx} \langmid
      \spobjset{\spc}{\spx}{\spexpr} \langmid
      \spobjapp{\spc}{\spf}{\spexpr}
    \\[1.5ex]
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptstr \langmid
      \sptobject \langmid
      \sptclass \langmid
  \\ & &
      \sptpydict{\sptype}{\sptype} \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptunion{\sptype, \ldots}
    \\[1.5ex]
    \sptenv & \defeq &
      \sptenvnil \langmid
      \sptvardef{\spx}{\sptype},~\sptenv \langmid
      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype}}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype}},~\sptenv
    \\[1.5ex]
    \spx, \spf & \defeq & \mbox{variable names}
    %% TODO should C be a claass name? ... should c be variable name?
  \end{langarray}

@@

  \caption{Surface syntax and types}
  \label{f:surface-types}
\end{figure}

@@

  \caption{Evaluation types, Surface-to-Evaluation mapping}
  \label{f:surface-to-eval-types}
\end{figure}

Evaluation types $\spteval$ are a simplification of the surface types.
These are the types that \SP{} promises to soundly enforce.
\Cref{f:surface-to-eval-types} presents the syntax of evaluation types
and a retraction $\mftypeF{\cdot}$ from surface types to evaluation types.
As the retraction shows, the evaluation types make two main changes:
\begin{enumerate}
  \item
    The parameterized type for Python dictionaries
    $\sptpydict{\sptype_0}{\sptype_1}$ gets replaced with
    a raw type $\sptrawpydict$.
    The raw type behaves the same as $\sptpydict{\sptdyn}{\sptdyn}$ would.
  \item
    Unions get replaced with the dynamic type, except for the special case
    of unions with none ($\sptoptional{\sptype_0}$).
\end{enumerate}

@@

\begin{table}[t]
  \begin{threeparttable}
  \centering
  \caption{Description of run-time checks for the evaluation types}
  \label{t:cast}

    \begin{tabular}{l@{\quad}l}
    {Eval. Type $\spteval$} & {Run-time check} \\\hline
    $\sptoptional{\spteval_0}$ & {Accepts either the none value or values that match $\spteval_0$} \\
    $\sptrawpydict$ & {Accepts any Python dictionary} \\
    $\sptchkdict{\spteval_0}{\spteval_1}$ & {Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
    $\sptdyn$ & {No check needed, accepts any value} \\
    $\sptobject$ & {Accepts any value\tnote{*}} \\
    $\sptclass_0$ & {Accepts instances of the class $\sptclass_0$} \\
    $\sptnone$ & {Accepts the Python none value} \\
    $\sptint$ & {Accepts integers} \\
    $\sptbool$ & {Accepts booleans (the integers \code{0} and \code{1})} \\
    $\sptstr$ & {Accepts strings}
    \end{tabular}

    \begin{tablenotes}
    \item {*} {\footnotesize In full \SP{}, the check for $\sptobject$ rejects primitive values~(\cref{s:c-types})}
    \end{tablenotes}
\end{threeparttable}
\end{table}

@@

    \inferrule*{
      \sptclassdef{\sptclass_0}{\sptclass_1}{\ldots}{\ldots} \in \sptenv_0
    }{
      \sptenv_0 \wtsub \sptclass_0 \spsubt \sptclass_1
    }

@@

  \caption{Expression typing, consistent subtyping, and materialization}
  \label{f:eval-types}
\end{figure}

@@

We end here, with materialization, rather than present a semantics
for the formalization.
After all, the main benefit of a full formal semantics is to validate
the behavior of complex expressions---and this job is
better left to the Redex mechanization, which at any rate supports a much
larger subset of Python~(\cref{s:surface}).

@@

%\subsubsection{Program Typing, Inheritance Restriction}
%
%\Cref{f:program-typing} presents the typing judgment for programs,
%which folds over a sequence of statements to build a type environment.
%The typing rules for variables and functions are standard.
%We draw attention to the rule for classes, in particular the override condition.
%A class may override fields and methods from its parent;
%if it does, the overriding definition must be a \emph{static subtype} of the original.
%For example, a method that returns an integer can be overridden by a method that returns
%a boolean, but it cannot be overriden by a method that returns the dynamic type.
%This restriction lets the \SP{} compiler avoid some run-time checks; it can assume
%that static methods return reliable values instead of having to insert checks for subtypes.
%
%\begin{figure}[t]
%  \fbox{$\sptenv \wtprog \spprog \dashv \sptenv$}
%  \begin{mathpar}
%    \inferrule*{
%      \sptenv_0 \wtprog \spstmt_0 \dashv \sptenv_1
%      \\\\
%      \sptenv_1 \wtprog \spprog_0 \dashv \sptenv_2
%    }{
%      \sptenv_0 \wtprog \spstmt_0, \spprog_0 \dashv \sptenv_2
%    }
%
%    \inferrule*{
%      \sptenv_0 \wtexpr \spexpr_0 : \spteval_0
%    }{
%      \sptenv_0 \wtprog \spvardef{\spx_0}{\spteval_0}{\spexpr_0} \dashv \sptvardef{\spx_0}{\spteval_0},~\sptenv_0
%    }
%
%    \inferrule*{
%      \sptenv_1 = \sptfundef{\spf_0}{\spteval_0}{\spteval_1},~\sptenv_0
%      \\\\
%      \sptenv_1 \wtexpr \spexpr_0 : \spteval_1
%    }{
%      \sptenv_0 \wtprog \spfundef{\spf_0}{\spann{\spx_0}{\spteval_0}}{\spteval_1}{\spexpr_0} \dashv \sptenv_1
%    }
%
%    \inferrule*{
%      \sptenv_1 = \sptclassdef{\spc_0}{\spc_1}{\sptvardef{\spx_0}{\spteval_0}}{\sptfundef{\spf_1}{\spc_0, \spteval_1}{\spteval_2}}, \sptenv_0
%      \\\\
%      %% TODO are field overrides exact?
%      \mbox{if $\spann{\spx_0}{\spteval_0'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\spteval_0 = \spteval_0'$}
%      \\\\
%      %% TODO can T override dyn?
%      \mbox{if $\sptfundef{\spf_1}{\spc_2, \spteval_1'}{\spteval_2'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\spteval_1' \spsubteq \spteval_1$ and $\spteval_2 \spsubteq \spteval_2'$}
%      \\\\
%      \sptenv_1 \wtexpr \spvardef{\spexpr_0} : \spteval_0
%      \\
%      \spann{\spself}{\spc_0}, \spann{\spx_1}{\spteval_1}, \sptenv_1 \wtexpr \spexpr_1 : \spteval_2
%    }{
%      \sptenv_0 \wtprog \spclassdef{\spc_0}{\spc_1}{\spvardef{\spx_0}{\spteval_0}{\spexpr_0}}{\spfundef{\spf_1}{\spself, \spann{\spx_1}{\spteval_1}}{\spteval_2}{\spexpr_1}} \dashv \sptenv_1
%    }
%  \end{mathpar}
%
%  \caption{Program typing}
%  \label{f:program-typing}
%\end{figure}
